---
title: "qgcomp vignette"
author: "Kelsey MacCuish"
output: pdf_document
---

Basic model of g-computation: 
 $$\mathbb{E}(Y^{\mathbf{X}_q} | \mathbf{Z,\psi,\eta}) = g(\psi_0 + \psi_1 S_q + \mathbf{\eta Z})$$
 - $g(\cdot)$: link function in GLM
 - $\psi_0$: model intercept
 - $\eta$: model coefficients for covariates 
 - $S_q$: index of joint value of exposures (all exposures being set to the same value)
 - $\mathbf{X}_q$: categorical bin of exposure
 - $\psi_1$: expected change in the outcome given a one quantile increase in all exposures simultaneously (overall mixture effect)
 
 
 $$[ \mathbb{E}(Y | \mathbf{X,\beta}) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3 ]$$
 - $\psi_1$ will equal the sum of the $\beta$ coefficients
 
 
```{r, load libraries}
library(qgcomp)
library(ggplot2)
library(dplyr)
library(broom)
```
 
 
```{r, load data}
data("metals", package = "qgcomp")
head(metals)
```
 
```{r, create linear model}
# save names of mixture variables in Xnm and covariated in covars
Xnm <- c('arsenic', 'barium', 'cadmium', 'calcium', 'chromium', 'copper',
         'iron', 'lead', 'magnesium', 'manganese', 'mercury', 'selenium',
         'silver', 'sodium', 'zinc')

covars <- c('nitrate', 'nitrite', 'sulfate', 'ph', 'total_alkalinity', 
            'total_hardness')

# run linear model
qc.fit <- qgcomp.glm.noboot(y~.,dat=metals[,c(Xnm, 'y')], family=gaussian())

# scaled coefficients/weights and statistical inference about mixture effect
# interpreted as the effect on the outcome of increasing every exposure by one quantile, possibly conditional on covariates (relative impact on outcome since all exposures are increasing by one quantile)

qc.fit
```
 
```{r, underlying model of qgcomp fit - linear model}
# quantized data
metals; quantile(metals$arsenic); qc.fit$qx

# how are the quantiles defined? 
qc.fit$qx %>% dplyr::select(arsenic_q) %>% group_by(arsenic_q) %>% count()
# 0-25%: 0 
# 25%-50%: 1
# 50%-75%: 2
# 75%-100%: 3

# underlying model of qgcomp fit
qc.fit$qx$y <- qc.fit$fit$data$y  # bring outcome back into quantized data

# adjusted coefficients from underlying model
# absolute change in outcome by increasing each exposure by 1
newfit <- lm(y ~ arsenic_q + barium_q + cadmium_q + calcium_q + 
               chromium_q + copper_q + iron_q + lead_q + magnesium_q + 
               manganese_q + mercury_q + selenium_q + silver_q + sodium_q + 
               zinc_q, data = qc.fit$qx)

sum(newfit$coefficients[-1]) # sums to psi_1 

# you can also calculate each scaled effect size by calculating a sum of the positive coefficients and a sum of the negative coefficients and calculating the corresponding proportion each exposure contributes to each sum
# e.g., calcium = 0.28/0.39 = 0.722 (0.722 is the scaled effect size calculated in qc.fit)

coef(qc.fit)
```
 
 
```{r, conditional odds ratio}
# conditional odds ratio
qc.fit2 <- qgcomp.glm.noboot(disease_state~., expnms = Xnm, 
                             data = metals[,c(Xnm, 'disease_state')], 
                             family = binomial(), q = 4)
qc.fit2

# underlying model of qgcomp fit
qc.fit2$qx$disease_state <- qc.fit2$fit$data$disease_state # bring outcome back into quantized data

# adjusted coefficients from underlying model
# absolute change in outcome by increasing each exposure by 1
newfit2 <- glm(disease_state ~ arsenic_q + barium_q + cadmium_q + calcium_q + 
               chromium_q + copper_q + iron_q + lead_q + magnesium_q + 
               manganese_q + mercury_q + selenium_q + silver_q + sodium_q + 
               zinc_q, data = qc.fit2$qx, family = binomial())

tidy(newfit2, conf.int = T)

sum(newfit2$coefficients[-1]) # sums to psi_1 

coef(qc.fit2)
```
 
 
```{r, marginal odds ratio}
# marginal odds ratio
qcboot.fit2 <- qgcomp.glm.boot(disease_state ~., expnms = Xnm, 
                               data = metals[,c(Xnm, 'disease_state')], 
                               family = binomial(),
                               q = 4, B = 10, # B should be 200-500+ in practice
                               seed = 125, rr = FALSE)
qcboot.fit2 # total scaled effect

# individual exposure beta coefficients and CIs
# (same as those from conditional odds ratio)
tidy(qcboot.fit2$fit, conf.int = T) 


# underlying model of qgcomp fit with bootstrapping
qcboot.fit2$qx$disease_state <- qcboot.fit2$fit$data$disease_state # bring outcome back into quantized data

# adjusted coefficients from underlying model
# absolute change in outcome by increasing each exposure by 1
newbootfit2 <- glm(disease_state ~ arsenic_q + barium_q + cadmium_q + calcium_q + 
               chromium_q + copper_q + iron_q + lead_q + magnesium_q + 
               manganese_q + mercury_q + selenium_q + silver_q + sodium_q + 
               zinc_q, data = qcboot.fit2$qx, family = binomial())

boot_func <- function(data, indices) {
  # Resample data
  boot_data <- data[indices, ]
  
  # Fit the glm model on the resampled data
  fit <- glm(disease_state ~ arsenic_q + barium_q + cadmium_q + calcium_q + 
               chromium_q + copper_q + iron_q + lead_q + magnesium_q + 
               manganese_q + mercury_q + selenium_q + silver_q + sodium_q + 
               zinc_q, data = boot_data, family = binomial())
  
  # Return the coefficients
  return(coef(fit))
}

set.seed(123)  # For reproducibility
boot_results <- boot(
  data = qcboot.fit2$qx,  # Your dataset
  statistic = boot_func,  # The bootstrap function
  R = 1000                # Number of bootstrap replicates
)

coef_names <- names(coef(newbootfit2))

# Calculate 95% bootstrap confidence intervals for each coefficient
ci <- t(apply(boot_results$t, 2, function(x) {
  quantile(x, probs = c(0.025, 0.975))
}))

# Add coefficient names for clarity
rownames(ci) <- coef_names
colnames(ci) <- c("Lower 95% CI", "Upper 95% CI")

# Mean of bootstrap coefficients
boot_means <- colMeans(boot_results$t)

# Combine results
bootstrap_summary <- data.frame(
  Coefficient = coef_names,
  Estimate = boot_means,
  `Lower 95% CI` = ci[, 1],
  `Upper 95% CI` = ci[, 2]
)

print(bootstrap_summary)
sum(bootstrap_summary$Estimate[-1])

sum(newbootfit2$coefficients[-1]) # sums to psi_1 

coef(qcboot.fit2)

```

 
```{r, age-adjusted linear model}
qc.fit3 <- qgcomp.glm.noboot(y ~ mage35 + arsenic + barium + cadmium + calcium + chloride + 
                           chromium + copper + iron + lead + magnesium + manganese + 
                           mercury + selenium + silver + sodium + zinc,
                           expnms = Xnm, 
                           metals, family = gaussian(), q = 4)
qc.fit3

# underlying model of qgcomp fit
qc.fit3$qx$y <- qc.fit3$fit$data$y # bring outcome back into quantized data
qc.fit3$qx$mage35 <- qc.fit3$fit$data$mage35 # bring maternal age back into quantized data
qc.fit3$qx$chloride <- qc.fit3$fit$data$chloride # bring chloride back into quantized data

# adjusted coefficients from underlying model
# absolute change in outcome by increasing each exposure by 1
newfit3 <- lm(y ~ mage35 +chloride + arsenic_q + barium_q + cadmium_q + calcium_q + 
               chromium_q + copper_q + iron_q + lead_q + magnesium_q + 
               manganese_q + mercury_q + selenium_q + silver_q + sodium_q + 
               zinc_q, data = qc.fit3$qx)

tidy(newfit3)

# add main exposure coefficients together - these should equal psi1 from qc.fit3
tidy(newfit3) %>% filter(!term %in% c("(Intercept)", "mage35", "chloride")) %>% summarize(sum = sum(estimate))

qc.fit3 # psi1 is equal to sum of coefficients above
```
 
```{r, plot of exposure effects}
plot(qc.fit3)
```
*weights = proportion of effect in a particular direction

*Left and right sides of the plot should not be compared because the lengths of the bars correspond to the effect size relative to the effects in the same direction
 
 
 
 

